\bf Title here

### 2020-07-27 paper/paper.tex
fix notation as per meeting 18

Derivative free methods provides another option to optimise the
objective function without evaluating any gradient or Hessian
information and a particular class of methods: direct search, has gained
its popularity through its conceptual simplicity. However, the whole
searching process in the algorithm remains a black-box. Plots are
usually used to evaluate and compare the performance of different
algorithms but it can easily become tedious because the code will have
to be modified significantly when comparing different parameters in the
algorithms. For example, a categorical variable with 5 levels can be
easily mapped onto color while mapping another categorical variable with
30 levels will not make the plot informative. Thus the plot needs to be
re-designed to better suits the characteristics of the parameter
(whether the parameter is a scalar or a matrix? whether the parameter is
quantitative or categorical? If categorical, how many levels does the
parameter have?). This motivates the design of a visual diagnostic
framework for optimisation algorithms based on the idea of a
\emph{global object}.

@@

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\) be a
data matrix, an n-d projection can be seen as a linear transformation
\(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis. Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or the index function and can be used to measure the
``interestingness'' of the projection. A number of indice functions have
been proposed in the literature including Legendre index
\citep{friedman1974projection}, hermite index
\citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}.

@@

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path} and guided tour uses the path that is geodesically the
shortest. Details of the mathematical construction of a tour path can be
found in \citet{buja2005computational}. Guided tour, along with other
types of tour, has been implemented in the \emph{tourr} package in R,
available on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

\hypertarget{notation}{%
\subsection{Notation}\label{notation}}

For a projection basis, let the three-subscript notation
\[\mathbf{A}_{jlk}\] represent the projection matrix in iteration \(j\)
with either a searching index \(l\) or an interpolation index \(k\). The
placeholder \(*\) will be the substitue for the non-applicable index. To
give two examples, \(\mathbf{A}_{12*}\) denotes the second searching
basis in the first iteration; \(\mathbf{A}_{1*2}\) denotes the second
interpolation basis in the first iteration.

If we are being pedantic, all the target bases, except the starting
basis, should have a searching index since it is the last candidate
basis that successfully allows the optimisation algorithm to end and
output a new basis for interpolation. If we denotes the length of each
search (number of basis searched in each iteration) as
\(l_1, l_2, \cdots, l_J\) and the length of each interpolation is
\(k_1, k_2, \cdots, k_J\) , then the searching index for target basis in
each iteration will become
\(\mathbf{A}_{1l_1*}, \mathbf{A}_{2l_2*}, \cdots, \mathbf{A}_{Jl_J*}\).
We simplify this notation for target bases by using a superscript as
\[\mathbf{A}^{0}, \mathbf{A}^{1}, \mathbf{A}^{2}, \cdots, \mathbf{A}^{J}\]
where \(\mathbf{A}^{0}\) is the starting basis. Notice that in the case
when the complex notation can't be avoided, we shall separate each index
using a comma and write the index in brackets if necessary. Using the
notation above, all the bases on the interpolation path (including
target and interpolation bases) can be written as in Equation
\ref{eq:interp-path}.

\begin{equation}
\{ 
\mathbf{A}^0, 
\mathbf{A}^1, 
\mathbf{A}_{1 \ast 1}, 
\mathbf{A}_{1 \ast 2},
\cdots, 
\mathbf{A}^2, 
\mathbf{A}_{2 \ast 1}, 
\mathbf{A}_{2 \ast 2}, 
\cdots, 
\mathbf{A}^J,
\mathbf{A}_{J \ast 1}, 
\mathbf{A}_{J \ast 2}, 
\cdots, 
\mathbf{A}_{J \ast K},
\}
\label{eq:interp-path}
\end{equation}

where \(K\) denotes the last interpolating index. A visual
representation of the interpolation path ( for iteration one) modified
from \citet{buja2005computational} is shown in Figure \ref{tour-path}.
If we use a dot to represent a plane in the searching space, the
searching points along with the current and target basis in iteration
one can sketched in Figure \ref{searching-points}. Combine the two
above, we can write all the bases in the first iteration (including the
starting basis) in its natural ordering as in Equation \ref{eq:vectorA}.

\begin{equation}
\{
\mathbf{A}^{0}, 
\mathbf{A}_{1 1\ast}, 
\mathbf{A}_{1 2\ast}, 
\cdots, 
\mathbf{A}^{1},
\mathbf{A}_{1\ast 1}, 
\mathbf{A}_{1\ast 2},  
\cdots, 
\mathbf{A}_{1\ast k_1}
\}
\label{eq:vectorA}
\end{equation}

\begin{figure}
\includegraphics[width=1\linewidth,height=0.6\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/tour_path_increasing} \caption{\label{tour-path}An illustration of the tour path}\label{fig:tour-path}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,height=0.6\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/tour_path_increasing} \caption{\label{searching-points}An illustration of the tour path}\label{fig:searching-points}
\end{figure}

@@

There are three existing methods for optimisating PPI function and we
review them below.

\citet{posse1995projection} proposed a stochastic direct search method,
a random search algorithm. Given a current basis \(\mathbf{A}^{j-1}\), a
candidate basis \(\mathbf{A}_{jl*}\) is sampled in the neighbourhood
defined by the radius of the p-dimensional sphere, \(\alpha\) of the
current basis \(\mathbf{A}^{j}\) by
\(\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}\). If
the candidate basis has a higher index value than the current basis, it
is outputed as the target basis \(\mathbf{A}^{j}\) along with other
metadata and the current iteration stops. If no basis is found to have
higher index value after the maximum number of tries \(l_{\max}\), the
algorithm stops with nothing outputted. \(c\) is the halfing parameter
and when \(c > 30\), the nieghbourhood parameter \(\alpha\) in the next
iteration will be reduced by half. The algorithm for a random search is
summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = \mathbf{A}^{j-1} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{j, l,*}$ is orthonormal\;
    Compute $I_{jl*}  = f(\mathbf{A}_{jl*})$\;
    \eIf{$I_{jl*} > I^{j-1}$}{
      $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

@@

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate ten bases in five random directions: $\mathbf{A}_{jl*}: \mathbf{A}_{j,(l + 9), *}$ within a small neighbourhood $\text{dist}$\;
    Find the direction with the largest index value: $\mathbf{A}_{j, l_{d}, *}$\;
    Construct the geodesic from $\mathbf{A}^{j-1}$ to $\mathbf{A}_{j, l_{d}, *}$\;
    Find $\mathbf{A}_{j,l_g, *}$ on the geodesic that has the largest index value \;
    Compute $I_{j,l_g, *} = f(\mathbf{A}_{j,l_g, *})$, $p_{\text{diff}} = (I_{j,l_g, *} - I^{j-1})/I_{j,l_g, *}$\;
      \If{$p_{\text{diff}} > 0.001$}{
        $A^{j} = A_{j,l_g, *}$, $I^{j} = I_{j,l_g, *}$\;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{log(l + 1)}\). The simulated annealing
algorithm works as follows. Given a neighbourhood parameter \(\alpha\)
and a randomly generated orthonormal basis \(A_{rand}\), a candidate
basis is constructed as
\(\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j} + \alpha \mathbf{A}_{rand}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis becomes the target basis. If it is
smaller, the candidate is accepted with probability
\(P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}\)
where \(I^{j-1}\) and \(I_{jl*}\) are the index value of the current and
candidate basis respectively. The algorithm can be summarised as in
Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}^{j-1}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{jl*} = (1 - \alpha)\mathbf{A}^{j-1} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{jl*}$ is orthonormal \;
    Commpute $I_{jl*} = f(\mathbf{A}_{jl*})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{jl*} > I^{j-1}$}{
        $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I^{j-1} - I_{jl*}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
          $A^{j} = A_{jl*}$, $I^{j} = I_{jl*}$\;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and
a \emph{global object} is constructed as the data frame supplied to the
visual diagnostic plots to better suit the characters of iterative
optimisation algorithms. Given an optimisation algorithm, two primary
variables of interest are the \emph{decision variable: \(\mathbf{A}\)}
and the \emph{value of the objective function: \(I\)}. To further
simplify the notation for \(\mathbf{A}\), we write the bases as a column
vector denoting by a single subscript by their natural ordering as in
Equation \ref{vector-basis}.

\begin{equation}
\{
\mathbf{A}_0, 
\mathbf{A}_1, 
\mathbf{A}_2, 
\cdots,
\mathbf{A}_T
\label{vector-basis}
\}
\end{equation}

where \(T\) is the total number of plane serached by the optimisation
algorithm including all the target planes, all the interpolating planes
and all the candidate planes but NOT the starting plane.

\emph{Iterators} indexes the data collected and has a time series
feature that prescribes the its natural ordering of the decision varible
in the searching. The simpliest iterator, \emph{index per observation:
\(t\)}, is a unique identifier for each observation in the data. For
each level of iteration, we design two types of indices: \emph{index per
iteration: \(j\)} is fixed for each observations in one iteration and
has an increment of one once a new iteration starts. \emph{index within
iteration: \(l\)} has an increment of one for each observation in an
iteartion and starts over from one once a new iteration starts. A sketch
of the difference between these three iterators is provided in Figure
\ref{iterators}. In projection pursuit optimisation algorithms, there is
one level of iteartion and hence exists three iterators: \texttt{id}
indices each observation by a unique number; \texttt{tries} is the index
per iteration iterator that gets updated once a search-and-interpolate
step is finished; and \texttt{loop} is the index within iteration
iterator and starts over from one at the beginning of a new
\texttt{tries}. We give the interpolating basis a different index \(k\)
for projection pursuit guided tour. It is similar in nature to \(l\) but
is specific for interpolating bases, which is usually not part of the
optimisation.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.25\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/iterators} \caption{\label{iterators} A sketch of the design of iterators in iterative algorithms.}\label{fig:iterators}
\end{figure}

There could exist other parameters that are also of our interest but
can't be classified as one of the three categories. They are defined
under the fourth category: \emph{other parameter of interest: \(S\)}. In
projection pursuit, three other parameters of our interest includes
\texttt{method}, \texttt{alpha} and \texttt{info}. \texttt{method}
identifies the name of the searching method used and we are interested
in comparing the performance between different algorithms under direct
search. The neighbourhood parameter \texttt{alpha} controls the size of
the sampling space and we are interested to understand how the searching
space shrinks as the algorithm progresses. \texttt{info} labels
different stages in the searching process. A sketch of the global object
for projection pursuit guided tour is presented in Figure
\ref{fig:glb-obj}. The full data structure can thus be shown as in
Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|cc|ccc|ccc}
t & \mathbf{A} & I & j &  l  & k & S_{1}& \cdots & S_{p}\\
\hline
1 & \mathbf{A}_0 & I_0 & 0 & \ast & \ast & \ast & \cdots & \ast\\
\hline
2 & \mathbf{A}_1 & I_1 & 1 & 1 & \ast & S_{1, 1} & \cdots & S_{1, p}\\
3 & \mathbf{A}_2 & I_2 & 1 & 2 & \ast & S_{2, 1} & \cdots & S_{2, p}\\
\vdots & \vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & \vdots & 1 & l_1 & \ast & \vdots &  \cdots & \vdots\\
\hline
\vdots &\vdots & \vdots & 1 & \ast & 1& \vdots &  \cdots & \vdots\\
\vdots &\vdots &\vdots& 1 & \ast & 2& \vdots& \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots &\vdots & 1 & \ast & k_1&\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
T+1 & \mathbf{A}_T & I_T & J &  l_J & \ast & S_{T, 1}& \cdots & S_{T, p}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

\begin{figure}
\includegraphics[width=1\linewidth,height=0.2\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/global_obj} \caption{\label{glb-obj}The global object in projection pursuit guided tour.}\label{fig:glb-obj}
\end{figure}

@@

We first simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution iwth zero mean and variance of one as
in equation \ref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distribution
with different weights and locations. Equation \ref{eq:sim-x2} to
\ref{eq:sim-x7} show the distribution of each variables and Figure
\ref{origin-data} shows the histogram of each variable except
\texttt{x3}. All the variables are then scaled to ensure the mixture has
variance of one.

@@

  \begin{itemize}
  \tightlist
  \item
    data: group by \texttt{tries} and filter the observations in the
    group that have count greater than \texttt{cutoff\ =\ 15}.
  \item
    x: \texttt{tries} is mapped to the x-axis
  \item
    y: the index value after statistical transformation
    \(y = \{Q_q(I_{jlk} \cdot \mathds{1} (j = x, k = \ast))\}\) is
    mapped to the y-axis where \(q\) takes one of \(0, 25, 50, 75, 100\)
    and \(Q_q(x)\) finds the qth-quantile for the vector \(x\).
    \(\mathds{1}(.)\) is the identity operator.
  \end{itemize}
\item
  Layer 2: point geom

  \begin{itemize}
  \tightlist
  \item
    data: group by \texttt{tries} and filter the observations in the
    group that have count less than \texttt{cutoff\ =\ 15}.
  \item
    x: \texttt{tries} is mapped to the x-axis
  \item
    y: \texttt{index\_val} is mapped to the y-axis
  \end{itemize}
\item
  Layer 3: line geom

  \begin{itemize}
  \tightlist
  \item
    data: filter the points with the highest index value in each
    \texttt{tries}
  \item
    x: \texttt{tries} is mapped to the x-axis
  \item
    y: \texttt{index\_val} is mapped to the y-axis
  \end{itemize}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    data: filter the observations with \texttt{info} being
    \texttt{interpolation} and mutate \texttt{id} to be the row number
    of the subsetted tibble
  \item
    x: \texttt{id} is mapped to the x-axis
  \item
    y: \texttt{index\_val} is mapped to the y-axis
  \end{itemize}
\item
  Layer 2: line geom

  \begin{itemize}
  \tightlist
  \item
    using line goemetry for the same data and aesthetics
  \end{itemize}
\end{itemize}

@@

We use the same dataset as the toy example above to explore the search
function \texttt{search\_better} and we want to learn how the index
value changes on the interpolation path for \texttt{holes} index. From
the left panel of Figure \ref{interruption}, we observe that when
interpolating from the current basis to the target basis, the index
value may not be monotone: we could reach a basis with higher index
value than the target basis on the interpolation path. In this sense, we
would be better off using the basis with the highest index value on the
interpolation path as the current basis for the next iteration (rather
than using the target basis).

Hence, an interruption is constructed to accept the interpolating bases
only up to the one with the largest index value. After implementing this
interruption, the search finds higher final index value with fewer steps
as shown in the right panel of Figure \ref{interruption}.

@@

  \begin{itemize}
  \tightlist
  \item
    data: filter the observations with \texttt{info} being
    \texttt{interpolation}; bind the global object from optimisation and
    interpolation and form polishing; mutate \texttt{id} to be the row
    number of the binded tibble.
  \item
    x: \texttt{id} is mapped to the x-axis
  \item
    y: \texttt{index\_val} is mapped to the y-axis
  \item
    color: \texttt{method} is mapped to the color aesthetic
  \end{itemize}
\item
  Layer 2: line geom

  \begin{itemize}
  \tightlist
  \item
    using line goemetry for the same data and aesthetics
  \end{itemize}
\end{itemize}

@@

>>>>

Derivative free methods provides another option to optimise the
objective function without evaluating any gradient or Hessian
information and a particular class of methods: direct search, has gained
its popularity through its conceptual simplicity. However, the whole
searching process in the algorithm remains a black-box. Plots are
usually used to evaluate and compare the performance of different
algorithms but it can easily become tedious because the code will have
to be modified significantly when comparing different parameters in the
algorithms. For example, a categorical variable with 5 levels can be
easily mapped onto colour while mapping another categorical variable
with 30 levels will not make the plot informative. Thus the plot needs
to be re-designed to better suits the characteristics of the parameter
(whether the parameter is a scalar or a matrix? whether the parameter is
quantitative or categorical? If categorical, how many levels does the
parameter have?). This motivates the design of a visual diagnostic
framework for optimisation algorithms based on the idea of a
\emph{global object}.

@@

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\) be a
data matrix, an n-d projection can be seen as a linear transformation
\(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis. Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or the index function and can be used to measure the
``interestingness'' of the projection. A number of index functions have
been proposed in the literature including Legendre index
\citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, natural Hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}.

@@

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path} and guided tour uses the path that is geodesically the
shortest. Details of the mathematical construction of a tour path can be
found in \citet{buja2005computational}. {[}Maybe add the sketch for
illustrating tour path.{]} Guided tour, along with other types of tour,
has been implemented in the \emph{tourr} package in R, available on the
Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.6\textheight]{/Users/hzha400/Documents/3.PhD/research/paper-tour-vis/figures/tour_path_increasing} \caption{\label{tour-path}An illustration of the tour path}\label{fig:tour-path}
\end{figure}

@@

There are three existing methods for optimising PPI function and we
review them below.

\citet{posse1995projection} proposed a stochastic direct search method,
a random search algorithm. Given a current basis
\(\mathbf{A}_{\text{cur}}\), a candidate basis \(\mathbf{A}_{t}\) is
sampled in the neighbourhood defined by the radius of the p-dimensional
sphere, \(\alpha\) of the current basis \(\mathbf{A}_{t}\) by
\(\mathbf{A}_{t} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\).
If the candidate basis has a higher index value than the current basis,
it is outputted as the target basis \(\mathbf{A}_{t}\) along with other
metadata and the current iteration stops. If no basis is found to have
higher index value after the maximum number of tries \(l_{\max}\), the
algorithm stops with nothing outputted. \(c\) is the halfing parameter
and when \(c > 30\), the neighbourhood parameter \(\alpha\) in the next
iteration will be reduced by half. The algorithm for a random search is
summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{t+1}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

@@

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{**}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate ten bases in five random directions: $\mathbf{A}_{l}: \mathbf{A}_{l+9}$ within a small neighbourhood $\text{dist}$\;
    Find the direction with the largest index value: $\mathbf{A}_{*}$ where $l < * < l+ 9$\;
    Construct the geodesic $\mathcal{G}$ from $\mathbf{A}_{\text{cur}}$ to $\mathbf{A}_{*}$\;
    Find $\mathbf{A}_{**}$ on the geodesic $\mathcal{G}$ that has the largest index value \;
    Compute $I_{**} = f(\mathbf{A}_{**})$, $p_{\text{diff}} = (I_{**} - I_{\text{cur}})/I_{**}$\;
      \If{$p_{\text{diff}} > 0.001$}{
         \KwRet{$\mathbf{A}_{**}$} \;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(t)\). Given an initial \(T_0\), the temperature at iteration \(t\)
is defined as \(T(t) = \frac{T_0}{log(t + 1)}\). The simulated annealing
algorithm works as follows. Given a neighbourhood parameter \(\alpha\)
and a randomly generated orthonormal basis \(A_{rand}\), a candidate
basis is constructed as
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis becomes the target basis. If it is
smaller, the candidate is accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{t+1} - I_{t}}{T(t)}\right],1\right\}\)
where \(I_{t}\) and \(I_{t+1}\) are the index value of the current and
candidate basis respectively. The algorithm can be summarised as in
Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Set $l = 0$\;
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

@@

Ggplot requires a data frame that contains all the variables to plot and
a \emph{global object} is constructed as the data frame supplied to the
visual diagnostic plots to better suit the characters of iterative
optimisation algorithms. Given an optimisation algorithm, two primary
variables of interest are the \emph{decision variable: \(\mathbf{A}\)}
and the \emph{value of the objective function: \(I\)}, which is the
projection basis and index value repsectively in projection pursuit.
\emph{State: \(S\)} is another variable of great importance since it
labels \(\mathbf{A}\) and \(I\) by the stage they are in the whole
search-and-interpolate process.

\emph{Iterators} indexes the data collected and prescribes the the
natural ordering of each observation in different levels. \emph{\(t\)}
is a unique identifier that has an increment of one for each observation
and \emph{\(j\)} is the counter for each search-and-interpolate round.
{[}add a bit more for \emph{\(l\)} after decide whether to combine \(l\)
and \(k\). {]}

There are other parameters that are of our interest and we denote them
as \emph{\(V_{p}\)}. In projection pursuit, this includes
\(V_1 = \text{method}\), denoting the name of the searching method used
and \(V_2 = \text{alpha}\), which is a neighbourhood parameter that
controls the size in sampling candidate bases. A sketch of the global
object for projection pursuit guided tour is presented in Figure
\ref{fig:glb-obj}. The full data structure can thus be shown as in
Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|ccc|ccc|ccc}
t & \mathbf{A} & I & S & j &  l  & k & V_{1}& \cdots & V_{p}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & \ast & V_{1, 1} & \cdots & V_{p, 1}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1 & \ast & V_{1, 2} & \cdots & V_{p, 2}\\
3 & \mathbf{A}_3 & I_3 & S_3 & 2 & 2 & \ast & V_{1, 3} & \cdots & V_{p, 3}\\
\vdots & \vdots &\vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_1 & \ast & \vdots &  \cdots & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2 & \ast & 1& \vdots &  \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & 2& \vdots& \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & k_1&\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_J & \ast & V_{1, T}& \cdots & V_{p, T}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

\begin{figure}
\includegraphics[width=1\linewidth,height=0.2\textheight]{/Users/hzha400/Documents/3.PhD/research/paper-tour-vis/figures/global_obj} \caption{\label{glb-obj}The global object in projection pursuit guided tour.}\label{fig:glb-obj}
\end{figure}

@@

We first simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution with zero mean and variance of one as
in equation \ref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distribution
with different weights and locations. Equation \ref{eq:sim-x2} to
\ref{eq:sim-x7} show the distribution of each variables and Figure
\ref{origin-data} shows the histogram of each variable except
\texttt{x3}. All the variables are then scaled to ensure the mixture has
variance of one.

@@

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the group that
    have count greater than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: the statistical transformed index value:
    \(Q_q(I_{T} \cdot \mathds{1} (S = \text{search}))\) is mapped to the
    y-axis where \(q\) takes one of \(0, 25, 50, 75, 100\) and
    \(Q_q(x)\) finds the qth-quantile for the vector \(x\).
    \(\mathds{1}(.)\) is the identity operator.
  \end{itemize}
\item
  Layer 2: point geom

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the group that
    have count less than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: \(I\) is mapped to the y-axis
  \end{itemize}
\item
  Layer 3: line geom

  \begin{itemize}
  \tightlist
  \item
    data: filter the points with the highest index value in group \(j\)
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: \(I\) is mapped to the y-axis
  \end{itemize}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    data: filter the observations with \(S = \text{interpolation}\) and
    mutate \(t\) to be the row number of the subsetted tibble
  \item
    x: \(t\) is mapped to the x-axis
  \item
    y: \(I\) is mapped to the y-axis
  \end{itemize}
\item
  Layer 2: line geom

  \begin{itemize}
  \tightlist
  \item
    using line geometry for the same data and aesthetics
  \end{itemize}
\end{itemize}

@@

We use the same dataset as the toy example above to explore the search
function \texttt{search\_better} and we want to learn how the index
value changes on the interpolation path for \texttt{holes} index. From
the left panel of Figure \ref{interruption}, we observe that when
interpolating from the current basis to the target basis, the index
value may not be monotone: we could reach a basis with higher index
value than the target basis on the interpolation path. In this sense, we
would be better off using the basis with the highest index value on the
interpolation path as the current basis for the next iteration (rather
than using the target basis). Hence, an interruption is constructed to
accept the interpolating bases only up to the one with the largest index
value. After implementing this interruption, the search finds higher
final index value with fewer steps as shown in the right panel of Figure
\ref{interruption}.

@@

  \begin{itemize}
  \tightlist
  \item
    data: filter the observations with \(S = \text{interpolation}\);
    bind the global object from optimisation and interpolation and form
    polishing; mutate \(t\) to be the row number of the binded tibble.
  \item
    x: \(t\) is mapped to the x-axis
  \item
    y: \(I\) is mapped to the y-axis
  \item
    colour: \(V\) is mapped to the colour aesthetic
  \end{itemize}
\item
  Layer 2: line geom
\end{itemize}

@@

{[}add example from geozoo{]}
